---
title: "Causal Inference Policy Evaluation Assignment 1"
output: html_document
date: "2025-03-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Define packages that you need
packages_vector <- c("tidyverse",    # package collection for data management 
                      "dplyr",       # set of functions to work with data 
                      "foreign",     # loading data (e.g. data from old Stata versions)
                      "haven",       # loading data (e.g. data from newer Stata versions)
                      "knitr",       # integrates computing and reporting
                      "fastDummies", # dummy creation
                      "sandwich",    # robust standard errors
                      "lmtest",      # robust standard errors
                      "jtools",      # summary statistics (robust standard errors)
                      "fBasics",     # summary statistics
                      "arsenal",     # summary statistics
                      "data.table",  # tables
                      "stargazer",   # tables
                      "xtable",      # tables
                      "expss",       # tables, labels 
                      "sjlabelled",  # labels
                      "lubridate",   # dates
                      "mfx",         # marginal effects for nonlinear models, e.g. probit 
                      "ggplot2")     # plots and graphs 
                     

# Uncomment the following line the first time you use these packages to install them
# install.packages(packages_vector)
lapply(packages_vector, require, character.only = TRUE) 
```


```{r}
setwd("C:/Users/schwdo/switchdrive/PhD Dominik/Courses/58920-01 – Causal Inference for Policy Evaluation/04_Assignments/Assignment_1")
# Loading given dataset
load("Assignment_1.RData")
```

```{r}
glimpse(raw.data)

```

## Question 1

Creating a dummy variable indicating if someone is under 40 years old

```{r}

raw.data[c("agegr_2", "agegr_3", "agegr_4")] %>% rowSums()

# Getting an idea of current variables
raw.data[c("agegr_2", "agegr_3", "agegr_4")] %>% head()

# Creating dummy variable for under 40 years of age (if not agegr_3 & agegr_4)
raw.data$under_40yo <- ifelse(raw.data$agegr_3 == 0 & raw.data$agegr_4 == 0, 1, 0)

raw.data[c("under_40yo", "agegr_2", "agegr_3", "agegr_4")] %>% head()
```

Compute the share of people below 40 by treatment status.
```{r}
raw.data %>%
  group_by(treat) %>%
  summarise(share_under_40 = mean(under_40yo, na.rm = TRUE))
```

```{r}
attach(raw.data)
```

Report the standardized bias and the difference in means along with a p-value.
```{r}
# Differences in characteristics and outcomes by treatment status
#-------------------------------------------------------------------

# a selection of controls
x_desc<- cbind(under_40yo)
x_desc_names<- colnames(x_desc)

# Define a function estimating the differences in variables across D
balance_check.model <- function(x){
  
  # Conditional means
  mean_d0 <- mean(x[treat==0])
  mean_d1 <- mean(x[treat==1])
  
  # Variances in subsamples
  var_d0 <- var(x[treat==0])
  var_d1 <- var(x[treat==1])
  
  # Difference in means
  diff_d <- lm(x ~ treat)
  cov <- vcovHC(diff_d, type = "HC")
  robust.se <- sqrt(diag(cov))
  
  # Absolute standardized bias (difference in means over average std dev - assuming similar size of the groups)
  sb <- abs((mean_d1-mean_d0)/sqrt((var_d0+var_d1)/2))*100
  
  # Store output as a list 
  list(mean_d0 = mean_d0, 
       mean_d1 = mean_d1,
       diff_d = diff_d$coefficients[2], 
       robust.se = robust.se[2], 
       pval = 2*pnorm(-abs(diff_d$coefficients[2]/robust.se[2])),
       SB = sb)             
}

# Apply to selection of covariates (MARGIN = 2 to loop over columns)
diff_output <- apply(X = x_desc, MARGIN = 2, FUN = balance_check.model)
# Convert output in list format into a data frame 
diff_output<-as.data.frame(rbindlist(diff_output))

# Add number of observations (don't forget the comma!)
raw.data$treat <- as.vector(raw.data$treat)
obs <- c(nrow(raw.data[raw.data$treat==0,]), 
         nrow(raw.data[raw.data$treat==1,]), 
         NA, NA, NA, NA)
diff_output <- rbind(diff_output, obs)

# Display in desired format
rownames(diff_output)<- c(x_desc_names, "Observations")
colnames(diff_output)<- c("E(X|D=0)", "E(X|D=1)", "Difference", "s.e.", 
                          "p-value", "Abs. SB")
print("Difference in means by treatment status and standardized bias")
print(round(diff_output,digits = 3))
```
Based on these metrics, briefly comment on the balancedness of the treatment
and control group with respect to age and whether we should account for age differences
in our analysis:
The dummy variable under_40yo seems well balanced between treatment and control groups, with low standardized bias (2.6%) and an insignificant difference in means at the 5% significance level).

We should still account for age differences in our analysis due to:
1. Heterogeneity: Age might moderate treatment effects (e.g., younger vs older participants respond differently), so including it enables subgroup or interaction analysis.
2. Robustness: Including covariates protects against chance imbalances or unobserved heterogeneity in smaller subgroups.
3. Improved precision: Even small imbalances, or natural outcome variation related to age, can lead to improved statistical efficiency when controlled for.

## Question 2
### (a)
```{r}

# Set maximum horizon to look at after start of unemployment, 24 months
maxdur <- 24

# Matrix with as many rows as observations, and as many columns as months in horizon 
emp <- matrix(0, nrow=nrow(raw.data), ncol=(maxdur))

# Name the columns for each month
emp_list <- paste("emp", 1:maxdur, sep="_")
colnames(emp) <- emp_list

# Replace i^th column = 1 if exited unemployment in month i
for (i in 1:maxdur) {
  emp[,i] = ifelse(raw.data$date_end < raw.data$date_start + 30*i & 
                   raw.data$date_end >= raw.data$date_start + 30*(i-1), 1, 0)
}

# Bind as data frame
emp <- as.data.frame(emp)
data_reg <- cbind(data_reg, emp)

```


## Question 3
```{r}

```

## Question 4
```{r}

```


## Question 5
```{r}

```


## Question 6
```{r}

```
