---
title: "Causal Inference: Policy Evaluation — Assignment 1"
author: "Your Name"
date: "2025-03-27"
output: 
  html_document:
    number_sections: true
    toc: true
    fig_caption: true
    keep_tex: true         # Optional: keeps LaTeX file for debugging
    latex_engine: xelatex  # Use if you're using custom fonts or want Unicode
    code_folding: true
fontsize: 11pt
mainfont: "Times New Roman"  # Optional: works with xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# Define packages that you need
packages_vector <- c("tidyverse",    # package collection for data management 
                      "dplyr",       # set of functions to work with data 
                      "foreign",     # loading data (e.g. data from old Stata versions)
                      "haven",       # loading data (e.g. data from newer Stata versions)
                      "knitr",       # integrates computing and reporting
                      "fastDummies", # dummy creation
                      "sandwich",    # robust standard errors
                      "lmtest",      # robust standard errors
                      "jtools",      # summary statistics (robust standard errors)
                      "fBasics",     # summary statistics
                      "arsenal",     # summary statistics
                      "data.table",  # tables
                      "stargazer",   # tables
                      "xtable",      # tables
                      "expss",       # tables, labels 
                      "sjlabelled",  # labels
                      "lubridate",   # dates
                      "mfx",         # marginal effects for nonlinear models, e.g. probit 
                      "ggplot2",
                     "causalweight",
                     "gt")     # plots and graphs 
                     
# Uncomment the following line the first time you use these packages to install them
# install.packages(packages_vector)
lapply(packages_vector, require, character.only = TRUE) 
```


```{r}
setwd("G:/My Drive/SwissTPH/Spring2025ClassesBasel/CausalInferencePolicyEvaluation/04_Small_Assignments/Assignment_1/")
# Loading given dataset
load("Assignment_1.RData")
```

```{r}
#glimpse(raw.data)
```

## Question 1

Creating a dummy variable indicating if someone is under 40 years old

```{r}

# Getting an idea of current variables
raw.data[c("agegr_2", "agegr_3", "agegr_4")] %>% head()

# Seeing if there are rows with 0s for each dummy
raw.data[c("agegr_2", "agegr_3", "agegr_4")] %>% rowSums()

# Finding rows which correspond to those under 40 (note all those under 18 and above 60 were dropped)
raw.data$under_40yo <- ifelse(rowSums(raw.data[c("agegr_3", "agegr_4")]) == 0, 1, 0)

# Getting an idea of current variables
raw.data[c("agegr_2", "agegr_3", "agegr_4", "under_40yo")] %>% head()
```

Examining the balance of this dummy variable across treatment groups
```{r}
cro(raw.data$treat, raw.data$under_40yo)
cro_cpct_responses(raw.data$under_40yo, raw.data$treat)
cro_cpct_responses(raw.data$treat, raw.data$under_40yo)
```

```{r}
attach(raw.data)
```


Calculating the standardize bias and difference in means
```{r}
balance_check.model <- function(x){
  
  # Conditional means
  mean_d0 <- mean(x[treat==0])
  mean_d1 <- mean(x[treat==1])
  
  # Variances in subsamples
  var_d0 <- var(x[treat==0])
  var_d1 <- var(x[treat==1])
  
  # Difference in means
  diff_d <- lm(x ~ treat)
  cov <- vcovHC(diff_d, type = "HC")
  robust.se <- sqrt(diag(cov))
  
  # Absolute standardized bias
  sb <- abs((mean_d1-mean_d0)/sqrt((var_d0+var_d1)/2))*100
  
  # Store output as a list 
  list(mean_d0 = mean_d0, 
       mean_d1 = mean_d1,
       diff_d = diff_d$coefficients[2], 
       robust.se = robust.se[2], 
       pval = 2*pnorm(-abs(diff_d$coefficients[2]/robust.se[2])),
       SB = sb)             
}

balance_check.model(raw.data$under_40yo)
```

The simple comparison of balance across groups reveals that the probability of being treated (i.e. being a participant) is comparable between those under the age of 40 and those 40 years old or older, with 45% of those under 40 being a participant and 46.3% of those 40 years old or older being a participant. The difference in means statistical test evaluated the null hypothesis that those under 40 year old contains have the same probability of being treated as those over 40 years old. The p-value of 0.073 suggests that is not enough evidence to reject the null, (assuming we decided on a 5% significance level a priori) i.e. there is not enough evidence to conclude that the two groups have a different probability of being treated.

I would say it is not so straightforward to decide whether or not it should be included. Strictly speaking, in terms of causal effect identification, it is not necessary to include if it does not influence the probability of being treated, even if it is associated with the outcome. However, including the variable could improve the precision of the estimate, especially if it is a strong predictor of the outcome. 

It is perhaps even more complicated because the question precisely asks if we should `account for age differences'. If this is taken to ask if we should account for age differences more generally, the answer might change. Here we have calculated an age variable that is dichotomizing a continuous (or discrete...) quantity. It could be the case that an age variable defined in a different manner could be predictive of treatment status, indicating that it should be included.

## Question 2
### (a)
```{r}
# Set the horizon
maxdur <- 24

# Create a matrix with n rows and 24 columns
emp <- matrix(0, nrow = nrow(raw.data), ncol = maxdur)

# Create column names like emp_1 to emp_24
emp_list <- paste("emp", 1:maxdur, sep = "_")
colnames(emp) <- emp_list

# Fill in the matrix
for (i in 1:maxdur) {
  emp[, i] <- ifelse(
    raw.data$date_end < raw.data$date_start + 30 * i,
    1,  # employed in month i
    0   # still unemployed in month i
  )
}

# Convert to data frame and combine with original data
emp <- as.data.frame(emp)
raw.data <- cbind(raw.data, emp)
```


```{r include=FALSE}
# Create the outcome variable - "monthly employment probability"

raw.data[c("id", "idobs", "date_start", "date_end", "treat", paste0("emp_", 1:24))]
raw.data$month_employ_prob <- raw.data %>% dplyr::select(paste0("emp_", 1:24)) %>% rowMeans()
```


```{r}
# Inspect distribution of monthly employment probability (for each spell)
hist(raw.data$month_employ_prob, main = "Monthly Employment Probability", xlab = "Probability")
hist(subset(raw.data$month_employ_prob, treat == 0), main = "Monthly Employment Probability - Untreated", xlab = "Probability")
hist(subset(raw.data$month_employ_prob, treat == 1), main = "Monthly Employment Probability - Treated", xlab = "Probability")
```

```{r}
# Sanity check for employment duration length
raw.data$duration <- raw.data$date_end - raw.data$date_start
hist(as.numeric(raw.data$duration), xlab = "days", main = "Length of Unemployment Spells")
hist(subset(as.numeric(raw.data$duration), treat == 0), xlab = "days", main = "Length of Unemployment Spells - Untreated")
hist(subset(as.numeric(raw.data$duration), treat == 1), xlab = "days", main = "Length of Unemployment Spells - Treated")
```


### (b)

```{r}
# Splitting data into under 40 years old and over 40 years old
data_u40 <- raw.data %>% filter(under_40yo == 1)
data_o40 <- raw.data %>% filter(under_40yo == 0)

treat_u40 <- data_u40$treat
treat_o40 <- data_o40$treat

x_u40 <- as.matrix(dplyr::select(data_u40, 
                              covs))
x_o40 <- as.matrix(dplyr::select(data_o40, 
                              covs))
x_names <- colnames(x_u40)

```


```{r include=FALSE}
# Estimate the p-score model for under 40
pscore.model_u40 <- glm(treat_u40 ~ x_u40, family = binomial(link = "probit"))
summ(pscore.model_u40, robust = "HC1")

# Estimate the p-score model for over 40
pscore.model_o40 <- glm(treat_o40 ~ x_o40, family = binomial(link = "probit"))
summ(pscore.model_o40, robust = "HC1")
```

```{r include=FALSE}
data_u40$pscore <- pscore.model_u40$fitted.values 
summary(data_u40$pscore)

data_o40$pscore <- pscore.model_o40$fitted.values 
summary(data_o40$pscore)

```

```{r}
# Create a factor variable for treatment status for the plot
data_u40$treat_f <- factor(treat_u40, 
                           levels = c(0,1), 
                           label = c("D=0", "D=1")) 

# Density plot for the propensity score by treatment status
ggplot(data_u40, aes(x = pscore, fill = treat_f)) + 
    geom_density(alpha=0.4) + 
    scale_fill_grey()+ 
    theme_bw(base_size = 20) +
    xlim(0, 1) + ggtitle("P-scores for under 40 years of age")


# Create a factor variable for treatment status for the plot
data_o40$treat_f <- factor(treat_o40, 
                           levels = c(0,1), 
                           label = c("D=0", "D=1")) 

# Density plot for the propensity score by treatment status
ggplot(data_o40, aes(x = pscore, fill = treat_f)) + 
    geom_density(alpha=0.4) + 
    scale_fill_grey()+ 
    theme_bw(base_size = 20) +
    xlim(0, 1) + ggtitle("P-scores for over 40 years of age")

```


```{r include=FALSE}
# As it is the ATE, the min and max pscores for both treated and not treated should be similar - i.e. "common support"
min(data_u40$pscore[data_u40$treat == 1])
max(data_u40$pscore[data_u40$treat == 1])

min(data_u40$pscore[data_u40$treat == 0])
max(data_u40$pscore[data_u40$treat == 0])

trim_pscores <- function(data) {
  # Compute min and max pscore for treated and control groups
  treated_min <- min(data$pscore[data$treat == 1], na.rm = TRUE)
  treated_max <- max(data$pscore[data$treat == 1], na.rm = TRUE)
  control_min <- min(data$pscore[data$treat == 0], na.rm = TRUE)
  control_max <- max(data$pscore[data$treat == 0], na.rm = TRUE)

  # Define common support
  min_overlap <- max(treated_min, control_min)
  max_overlap <- min(treated_max, control_max)
  
  # Trim the dataset
  trimmed_data <- data[data$pscore >= min_overlap & data$pscore <= max_overlap, ]

  return(trimmed_data)
}

data_u40_trimmed <- trim_pscores(data_u40)
data_o40_trimmed <- trim_pscores(data_o40)
```


```{r}
# Only 1 observation removed from each
nrow(data_u40) - nrow(data_u40_trimmed)
nrow(data_o40) - nrow(data_o40_trimmed)

```

```{r}
# Checking for aligment
min(data_u40_trimmed$pscore[data_u40_trimmed$treat == 1])
max(data_u40_trimmed$pscore[data_u40_trimmed$treat == 1])

min(data_u40_trimmed$pscore[data_u40_trimmed$treat == 0])
max(data_u40_trimmed$pscore[data_u40_trimmed$treat == 0])

# Density plot for the propensity score by treatment status
ggplot(data_u40_trimmed, aes(x = pscore, fill = treat_f)) + 
    geom_density(alpha=0.4) + 
    scale_fill_grey()+ 
    theme_bw(base_size = 20) +
    xlim(0, 1) + ggtitle("P-scores for under 40 years of age - Trimmed")

# Density plot for the propensity score by treatment status
ggplot(data_o40_trimmed, aes(x = pscore, fill = treat_f)) + 
    geom_density(alpha=0.4) + 
    scale_fill_grey()+ 
    theme_bw(base_size = 20) +
    xlim(0, 1) + ggtitle("P-scores for over 40 years of age - Trimmed")
```
### (c)
```{r}
# Define the outcome - "monthly employment probability"
y1_u40 <- data_u40_trimmed$month_employ_prob 
y1_o40 <- data_o40_trimmed$month_employ_prob
x_u40 <- as.matrix(data_u40_trimmed[covs])
x_o40 <- as.matrix(data_o40_trimmed[covs])
treat_u40 <- data_u40_trimmed$treat
treat_o40 <- data_o40_trimmed$treat

x_u40

```

```{r include=FALSE}
s_boot <- 10
length(y1_u40) == length(treat_u40) 
length(y1_u40) == dim(x_u40)[1]

length(y1_o40) == length(treat_o40)
length(y1_o40) == dim(x_o40)[[1]]

# # Estimating the ATE 
# # IPW - Estimating the ATET based on the causalweight package
# ipw_ate_u40 <- treatweight(y = y1_u40, # take initial data 
#                         d = treat_u40, 
#                         x = x_u40, 
#                         ATET = FALSE, # if = FALSE, estimates ATE (default)
#                         trim = 0.0, # data was already trimmed previously
#                         boot = s_boot) # number of bootstrap replications
# 
# ipw_ate_o40 <- treatweight(y = y1_o40, # take initial data 
#                         d = treat_o40, 
#                         x = x_o40, 
#                         ATET = FALSE, # if = FALSE, estimates ATE (default)
#                         trim = 0.0, # data was already trimmed previously
#                         boot = s_boot) # number of bootstrap replications
```


```{r}
# Create function to do IPW for each month
reg_monthly_ipw <- function(y, d, x) {
  ipw <- treatweight(y = y, d = d, x = x, ATET = FALSE, trim = 0, boot = s_boot)
  list(effect = ipw$effect, se = ipw$se)
}
```



```{r}
### For over 40 years of age

# Create a matrix for employment status for each month
emp_o40 <- data_o40_trimmed %>% dplyr::select(colnames(emp)) %>% as.matrix()

# Apply IPW each month
ipw_monthly_o40 <- apply(emp_o40, 2, function(y) reg_monthly_ipw(y, treat_o40, x_o40))

# Look at results
ipw_monthly_o40_df <- rbindlist(ipw_monthly_o40)
ipw_monthly_o40_df$month <- 1:maxdur

# Add confidence intervals
ipw_monthly_o40_df <- ipw_monthly_o40_df %>%
  mutate(
    cil = effect - 1.96 * se,
    cih = effect + 1.96 * se,
    sig = ifelse(abs(effect / se) > 1.64, effect, NA)
  )
```


```{r}
### For under 40 years of age

# Create a matrix for employment status for each month
emp_u40 <- data_u40_trimmed %>% dplyr::select(colnames(emp)) %>% as.matrix()

# Apply IPW each month
ipw_monthly_u40 <- apply(emp_u40, 2, function(y) reg_monthly_ipw(y, treat_u40, x_u40))

# Look at results
ipw_monthly_u40_df <- rbindlist(ipw_monthly_u40)
ipw_monthly_u40_df$month <- 1:maxdur

# Add confidence intervals
ipw_monthly_u40_df <- ipw_monthly_u40_df %>%
  mutate(
    cil = effect - 1.96 * se,
    cih = effect + 1.96 * se,
    sig = ifelse(abs(effect / se) > 1.64, effect, NA)
  )
```

```{r}
ipw_monthly_u40
```


```{r}
# Create tables by group
ipw_monthly_u40_gt <- ipw_monthly_df %>%
  filter(Group == "Under 40") %>%
  gt() %>%
  tab_header(
    title = "Monthly IPW Estimates: Under 40",
    subtitle = "Inverse Probability Weighting (ATE Estimates per Month)"
  ) %>%
  fmt_number(columns = c(effect, se, cil, cih, sig), decimals = 3) %>%
  cols_label(
    month = "Month",
    effect = "ATE",
    se = "Std. Error",
    cil = "CI Lower",
    cih = "CI Upper",
    sig = "Significant Effect"
  )

ipw_monthly_o40_gt <- ipw_monthly_df %>%
  filter(Group == "Over 40") %>%
  gt() %>%
  tab_header(
    title = "Monthly IPW Estimates: Over 40",
    subtitle = "Inverse Probability Weighting (ATE Estimates per Month)"
  ) %>%
  fmt_number(columns = c(effect, se, cil, cih, sig), decimals = 3) %>%
  cols_label(
    month = "Month",
    effect = "ATE",
    se = "Std. Error",
    cil = "CI Lower",
    cih = "CI Upper",
    sig = "Significant Effect"
  )

ipw_monthly_u40_gt
ipw_monthly_o40_gt
```


It seems that the average treatment effect is quite different for those over 40 years old and those under 40 years old, with an average treatment effect of 0.135 and 0.191, respectively. This suggests that the training program had a greater impact on the older age group.

## Question 3


```{r}
# Plotting results by month by age group
ipw_monthly_u40_df$Group <- "Under 40"
ipw_monthly_o40_df$Group <- "Over 40"
ipw_monthly_all <- bind_rows(ipw_monthly_u40_df, ipw_monthly_o40_df)


ggplot(ipw_monthly_all, aes(x = month, y = effect, color = Group, fill = Group)) +
  geom_line(size = 1) +
  geom_ribbon(aes(ymin = cil, ymax = cih), alpha = 0.2, color = NA) +
  geom_point(aes(y = sig), shape = 18, size = 3) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_bw(base_size = 18) +
  labs(
    title = "Monthly ATEs on Employment Probability by Age Group",
    x = "Months after program start",
    y = "Average Treatment Effect"
  )
```

The figure would suggest that effect of the program on employment probability increases for both age groups at a similar rate over time up until month five. After month 5, the program has a greater effect for those over 40 years old than those under 40 years old. It is also worth noting that the average treatment effect is positive for each month, indicating that, for each month, the group of program participants had an average employment probability higher than non-participants. 

## Question 4

See other document
